Honeybees vary collective decision making in different landscapes

### 2022-08-26 inst/publication/Manuscript/paper.Rmd
updated MS

---
title: "Honey bees vary collective decision making in different landscapes"
# author: "Joseph Palmer, Ash Samuelson, Rich Gill, Elli Leadbeater and Vincent A. A. Jansen"
output:
  word_document
always_allow_html: true
bibliography: reffs.bib
csl: science.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
devtools::load_all()
library(showtext)
showtext_auto()

@@

Honeybee foraging is an extraordinary collective behaviour which emerges from one of the most sophisticated communication systems in the animal kingdom: the waggle dance. Through a series of feedback loops, this elaborate behaviour -unparalleled in even a single other social insect- allows colonies to decide where to forage without requiring any individual to directly compare resources. This behaviour is a textbook example of an emergent collective process, yet a growing body of evidence now shows that using the dance does not always improve foraging success and sometimes not having access to the dance even increases it. Although the value of dancing varies across environments, it is not known if and to what extent colonies can adjust dance use in response. Here, develop a method to quantify dance use and show that colonies vary the use of the waggle dance in different landscapes, with some colonies relying on dance use almost exclusively and others not at all. Further probing of the drivers of this variation in dance use found a colonies use of the dance correlates with different landscape characteristics, suggesting colonies adjust collective decision-making in response to environmental conditions. Our results provide new evidence of flexible dance use in honey bee colonies and our methodology outlines a means to identify the ecological conditions in which honeybee colonies rely on dance communication, opening the door to large-scale experimental exploration of the selection pressures that may have driven the evolution of this remarkable collective behaviour.

@@

In group living animals, many decisions are taken collectively by integrating information from multiple individuals to produce a social behaviour which extends beyond that of the individual units [@Sumpter2006]. Such systems are self-organised, relying on the constituents using socially provided information and following behavioural 'rules' to produce an emergent collective outcome [@Couzin2003;@Bonabeau1997]. The behavioural architectures which produce these emergent collective behaviours have in some cases, such as in social insects, grown to dazzling levels of complexity. However, this complexity is not always beneficial. The benefits of working collaboratively versus independently vary across environmental contexts and individuals often adjust their own use of information accordingly [@Dechaume-Moncharmont2005;@Beekman2008]. Although the benefits of collective behaviours are known to vary across ecological conditions, in many cases we have a more limited understanding of if and how a group can flexibly use collective information to make use of the most profitable strategy under different conditions.

The extraordinary waggle dance communication system of the honey bee (*Apis* spp.) is a key example of a behavioural mechanism which produces emergent collective decisions [@Seeley1995]. Within colonies, a series of simple rules which determine when and how much bees perform dances (Fig. 1) mean that choices between feeding sites are made by the group rather than individuals [@Seeley1991;@Seeley1994A;@Hasenjager2022]. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the richer of two equidistant sources [@Seeley1991], or the closer of two equally rich sources [@Hasenjager2022]. This mechanism allows a colony to pool individual assessments of the landscape and processes them collectively to identify the most profitable resources to focus on, all without the need for any individual to compare resources.

```{r, fig.cap="Figure 1. **The honeybee waggle dance carries information about the location of a resource.** The duration of the waggle run indicates the distance to the resource and the angle of the dance relative to the vertical indicates the direction of the resource (circle in right panel), relative to the direction of the sun [@Frisch1993]. Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Overall resource “quality” -the net energetic gain of a foraging trip- is provided through the number of waggle runs performed [@Frisch1993;@Boch1956;@Esch1961;@Seeley1994;@Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A]."}
knitr::include_graphics("../Results/figures/manuscript_figures/fig1.png")
```

Despite it's complexity, research has shown that the ***value*** the waggle dance provides to foraging success varies according to environmental context and colonies without access to dance information achieve equal or even greater foraging success than their wild-type counterparts [@Sherman2002;@Dornhaus2004;@Gruter2013;@Price2019]. Interestingly, studies have shown that individuals often ignore the dance when looking for forage sites [@Gruter2009]. Whilst it likely that individual variation in dance use translates into variations in collective decision making at the colony level [@Donaldson-Matasci2012;@Price2019], this has yet to be empirically established. Identifying if and how colonies vary waggle dance ***use*** is not only important for deciphering how honey bees collectively respond to changes in environmental conditions [@Gruter2014], relating this to the environmental context of the behaviour is key to illuminating its ultimate evolutionary drivers [@Price2015].

Quantifying dance use within colonies requires measuring dance use across the foraging workforce [@Seeley1995]. This has only previously been quantifiable by documenting foragers' individual search history, which requires tracking individual bees and is time -and labour- intensive [@Seeley1995]. Individual based models, which explore the collective effects of individuals following different behaviours, provide an insight into how individual information use scales to the level of the group [@Dornhaus2006;@Beekman2008], however, these are difficult to relate to empirical observations and so are limited to showing what a colony under certain conditions might do, rather than quantifying what they are doing. Empirically determining how dance information is used at a colony level requires new methods which can identify the behaviours of individual foragers and integrate them with the mechanisms by which the dance functions to make collective decisions.

Here, we derive a mathematical model to quantify waggle dance use in colony foraging by identifying the proportion of foraging trips that are made by bees recruited through waggle dances. To validate our model, we fit it to waggle dance observations collected from 20 sites across and agri-rural and urban environment (10 in each). We then use our model to explore how variations in landscape structure influences dance use in colony foraging.

@@

The patterns of waggle dance-encoded foraging distances might differ under scenarios where colonies rely on individual search versus recruitment. To test this, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor [@Seeley1992] (Fig. 2a,b, details in Materials and Methods). As it is known in the simulation which individuals in the hive forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor by each type of forager. Fig 2c shows that the shapes of the resource distance distributions for bees engaging in the two types of foraging trips are different. The distance distribution for the scout trips is akin close to that of an exponential distribution (Fig. 2c), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see Materials and Methods). The distribution of the distances reported for recruit trips (Fig. 2c) is a Rayleigh distribution which is the nearest-neighbour distribution in a two-dimensional environment [@Pyke1978] (see Materials and Methods).

```{r, fig.cap="Figure 2. **Simulating honey bee foraging**. In our simulation model with scouting only (A), foragers leave the hive on a search path (white lines) and continue until they encounter a resource (circles, colours indicate different resource quality). When foraging with recruitment (B) foragers continue to identify resources in scouting trips (white lines) and convey this information on the dance floor (brown disc) where foragers can sample dances reporting on scouting and recruiting trips and follow these directions (yellow lines). (C) Complementary cumulative frequencies of foraging distances reported from scouting and recruit trips. Note the difference in the shape of the distributions. The scout distribution is best fit by an exponential (grey fit line), the recruit distribution by a Rayleigh distribution (black line)"}
knitr::include_graphics("../Results/figures/manuscript_figures/fig2.png")
```

@@

We constructed two models of foraging: an individual model describing a hypothetical colony that relies only on individual search to find foraging sites, and a collective model which describes a colony using waggle dance communication to exploit resources. In the individual model, all trips are "scout" trips, but as recruitment becomes more important, the proportion of “recruit” trips will increase. By setting $p = 1$, we obtain a model based on the sole use of individual search (scouts only) and, alternatively, by allowing the proportion of scout trips, $p$, to take on any value between 0 and 1, we can model the extent to which foraging occurs collectively. (Fig 2). In fitting these two models to distribution of waggle run durations decoded from real honeybee colonies we are thus able, using model selection [@Burnham2002], to infer if and, by estimating the parameter $p$, to what extent honeybee colonies use waggle dance information when foraging.

@@

```{r, fig.cap="Figure 4. **The honey bee foraging model fitted to data from 20 hives**. (A) Location of study hives in Southern England, shaded area in the main plot indicates Greater London. For 16 hives for the collective foraging model provided best explanation (black circles) or for 4 hives the individual search model provided the best explanation (red triangles) as indicated by lowest AIC score. (B). Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. The p-value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data, the red dashed line marks the significance threshold of 0.05. For values exceeding the threshold there is no statistically significant difference between the model and the data, indicating the model provides a good fit. For the hive in (C) the individual model (blue line) provided a better fit than the collective foraging model (red line). For the hive in (D) the collective foraging model (red line) provided a better fit than the individual model (blue line). The typical “hump” in the distribution in (D) which is indicative of contribution of recruitment dances (compare to Fig 3H). Panels show the compliment cumulative frequencies with binned frequency distributions as inset."}
knitr::include_graphics("../Results/figures/manuscript_figures/fig4.png")
```

Our results show that, whilst colony-level foraging is mostly comprised of a mixture of scout and recruit foraging trips, in some circumstances, colony foraging can be better described by individual foraging alone. Thus, in some environments, the majority of foraging trips involve scouting to find new food sites rather than recruitment through dances. Note that this does not imply that these bees do not engage in dance following, because bees regularly follow dances but choose not to visit the advertised site [@Gruter2011], although it has also been shown that bees may cease dance following if it is proving unproductive [@Gruter2011;@Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individual's foraging behaviour over a landscape. As our results show that our model of individual foraging provides a good fit to colony foraging in 11 different sites, and provides a more parsimonious description of foraging than a model of collective foraging in 4 sites, these findings support the idea that individual decision making can dominate colony foraging and demonstrates further evidence for flexible waggle dance use by honey bee colonies.

quantifying the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of recruit trips, $1-p$, for each site. Since our sites varied in land-use characteristics, and thus potentially forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see Materials and Methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see Materials and Methods) to determine the principal components that represent combinations of land-use types which explained the most variation in waggle dance use within agri-rural and urban environments. As for one of our sites neither model provided a plausible description, this site was removed from the PLS analysis.

Waggle dance use varied widely regardless of environment, ranging from 0 to 1 in both the urban and agri-rural environments whilst showing a significant, positive interaction with waggle dance use. In the agri-rural sites, a single principle component explained ~73% of the variation in waggle dance use (beta regression: $R^2 = 0.73$, $\phi$ = 4.9, p < 0.05, Fig. 5A). This finding is also shown in the urban environment which also shows a single principle component explained ~73% of the variation in waggle dance use (beta regression: $R^2$ = 0.73, $\phi$ = 10.4, p < 0.05, Fig. 5C). Probing the individual loadings of these principal components allows us to evaluate the driving mechanisms in more detail. As our sample size is limited  (10 urban and 9 agri-rural sites), we used jackknife resampling to evaluate the robustness of our results to influential points (see Materials and Methods, Supplementary Material) and evaluate how these land-use types contribute to the principle component which significantly correlates with waggle dance use.

@@

In the agri-rural environments the first principal component correlates positively with arable land (29% of land coverage; Table 1) and negatively with built-up areas (17% of land coverage); note that it also correlates negatively with non-agricultural unmanaged green space and water, but together these represent less than 3% of land-use (Table 1; Fig. 5B). These land-use types maintain a significant correlation with the first principal component over the jackknifed PLS (Fig. 5B), with the exception of non-agricultural unmanaged green space which sits on the boarder, indicating the results are robust. As arable land increases whilst built-up areas decrease, the proportion of trips that are driven by individual search increases. Arable land in the UK is typically considered nutritionally poor for bees (note that oilseed rape fields were not included within this category), while there is evidence to suggest that the residential areas that were captured within the “built-up” category are forage-rich hotspots, typically supporting relatively high bee diversity and abundance within gardens [@Baldock2019].

In the urban environment, our findings were more equivocal. The first principal component correlated negatively with the dominant land-use type by coverage, sparse residential (land coverage ~35%; Table 1), as would be expected since such land includes a high proportion of gardens. However, this relationship is very sensitive to site removal through the jackknife sampling (Supplementary Fig. 2.), and is not a significant contributor to the variance in the loadings (Fig. 5D.). The first principal component also correlates positively with railways but these constitute ~1% of land cover (Table 1). Furthermore, the first principal component correlates negatively with parks, allotments and cemeteries which would typically be considered forage-rich for honeybees [@Baldock2019] (although note that the amount of land cover for this component is limited and the jackknifed PLS identified substantial variation in the loadings identified for the first principal components; Supplementary Fig. 3.).

@@

Understanding how individual variation in information use scales to produce flexible collective processes at the group level is a key challenge in behavioural ecology. Here, we have presented a method to quantify information use at a collective level in a model social insect. After verifying that this method accurately captures empirical patterns of collective and individual foraging, we have shown how it can be used to evaluate how environmental factors shape collective foraging at the colony level. Whilst recruitment is well known to occur through other mechanisms besides the waggle dance, such as through olfactory cues [@Arenas2007], the exceptionally close fit of our model to waggle run durations underlines the importance of the waggle dance in honey bee foraging. Further examining our estimates of dance use show that a colonies use of the waggle dance varies across environments and does not appear to be randomly determined.

Within both the agri-rural and urban environments, dance use had a significant interaction with land-use type, ranging from total dance use to complete reliance on individual information. The notion that individual variation in information use influences the collective effect has only previously been inferred theoretically. Our findings demonstrate this empirically. Individual-based models [@Dornhaus2006;@Beekman2008;@Price2019] have outlined how variations at an individual level could drive group level dynamics, however, without verifying this upon empirical data the insights remain theoretical. Here, we have shown how an individual based model can be used to better understand a system in a way which allows us to build more informative models which can be empirically verified. Not only do our findings add support for our understanding of how the dance functions to make collective decisions [@Seeley1995], but they also provide a new dimension to our understanding of how variation at an individual level influences the collective level.

Whilst the benefits of dance communication are known to vary across environments [@Donaldson-Matasci2012;@Gruter2011;@Schurch2014;@Price2019], our results show that colonies vary their use in response, affording the colony a flexible dance use strategy which they can tailor to different conditions to maximise foraging success. Identifying the circumstances in which dance communication is important have been hampered by the requirement to monitor real-world colony weight [@Sherman2002;@Dornhaus2004;Gruter2013;@Price2019;@Dornhaus2006;@Nurnberger2017;@Kirchner1998] -a noisy proxy of foraging success that is influenced by many other abiotic and biotic factors, including colony health- over ecologically credible time periods [@Schurch2014]. Doing this at a scale that allows inter-colony variation in foraging environments at sufficient replication is a major logistical hurdle (but see [@Nurnberger2017]), particularly given that multiple landscape variables may interact to determine the utility of dance communication. Our method circumvents this issue by providing a time and labour efficient methodology to quantify this key collective behaviour and map the environment along the major axis of honeybee information use.

Examining the land-use types which contribute most to the variation in waggle dance use in each of our study environments reveals insights into the drivers of waggle dance use. Our agri-rural results suggest a potential decrease in reliance on waggle dance recruitment as resources become harder to find and foraging trips become longer [@Samuelson2021], broadly agreeing with the findings from other theoretical [@Dornhaus2006] (11) and empirical [@Price2019] studies which examined foraging performance with and without the dance. These results suggest that, even though collective foraging is not always beneficial, in environments where high-quality resources are present but relatively scarce, collective foraging is being carried out by colonies and under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the waggle dance use changes with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5A and 5C). These findings illustrate how our model can be used to estimate reliance on collective behaviour within different landscapes, for exploration within datasets where land-use is systematically chosen by design.

Although the number of colonies used in this analysis are comparable to similar studies evaluating collective foraging in honey bees [@Sherman2002;@Dornhaus2004;@Price2019;@Nurnberger2017;@Kirchner1998],  probing the landscape drivers of colony level waggle dance use in more detail requires more colony level data. As longer term effects, such as resource stability [@Schurch2014], may also influence waggle dance use, these data need to be gathered over sufficient time frames to ensure these effects are accounted for. Recently, technological advances have emerged which enable colony metrics to be collected faster, more accurately and over greater time spans than could be gathered by hand, allowing individuals to be tracked within colonies and theories of individual behaviour to be evaluated in more depth than could have been done previously [@Wild2021]. Here, we have presented an accurate mathematical description of colony foraging which extends our ability to quantify collective behaviour across environments. With the advances in the decoding of the waggle dance through automated methods [@Wario2017], we face the prospect of waggle dance data becoming “big data”. Our methodology thus provides a means of analysing such large data sets and provide a framework to start exploring how plasticity in collective decision making can help to buffer the effects of landscape change, and in this case, how flexibility in the use of the waggle dance may help to exploit resources when experiencing novel landscape mosaics, such as in highly urbanized conurbations.

@@

All of the data and code required to reproduce this article is contained as within the 'wagglefit' which can be found at (link to finalised Zenodo archive finalised)

https://github.com/joseph-palmer/wagglefit

>>>>

---
title: "Honeybees vary collective decision making in different landscapes"
# author: "Joseph Palmer, Ash Samuelson, Elli Leadbeater, Richard Gill, and Vincent A. A. Jansen"
output:
  word_document
always_allow_html: true
bibliography: reffs.bib
csl: science.csl
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
devtools::load_all()
library(showtext)
showtext_auto()

@@

Honeybee (*Apis mellifera*) colony foraging decisions arise from the waggle dances of individual foragers, refined and filtered through a series of feedback loops that produce emergent collective behaviour. This process is a key example of animal communication at the height of eusociality, but a growing body of evidence suggests that its value for colony foraging success is heavily dependent on local ecology. Here, we develop a method to quantify the extent to which a colony forages collectively based on dance-decoding and show how it can be used to relate variation in collective foraging to land-use. We show that the extent to which colonies forage collectively varies across landscape types, such that it is sometimes almost entirely dominant but at other times almost negligible. By providing a means to identify the ecological conditions in which waggle dances are important on a large scale and without experimental manipulation, our methodology opens the door to exploration of the selection pressures that may have driven the evolution of this remarkable group behaviour.

@@

In group living animals, many decisions are taken collectively by integrating information from multiple individuals to produce group behaviour that extends beyond that of the individual units [@Sumpter2006]. Such systems are self-organised, relying on the use of simple behavioural 'rules' that filter social information to produce an emergent collective outcome [@Couzin2003;@Bonabeau1997]. In honeybee colonies, and other highly eusocial insects, the behavioural architectures that produce these emergent behaviours have become particularly complex. Honeybee colony foraging is driven by the extraordinary waggle dances of individual foragers, which famously communicate food source locations. A series of rules that determine when and how much bees dance (Fig. 1) mean that choices between feeding sites are made by the group [@Seeley1991;@Seeley1994A;@Hasenjager2022]. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the richer of two equidistant sources [@Seeley1991], or the closer of two equally rich sources [@Hasenjager2022]. This mechanism allows a colony to collectively allocate foraging effort adaptively without the need for any individual to compare resources

```{r, fig.cap="Figure 1. **The honeybee waggle dance carries information about the location of a resource.** The angle of the dance relative to the vertical indicates the direction of the resource (circle in right panel), relative to the direction of the sun [@Frisch1993], the duration of the waggle run indicates the distance to the resource. Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Overall resource “quality” -the net energetic gain of a foraging trip- is provided through the number of waggle runs performed [@Frisch1993;@Boch1956;@Esch1961;@Seeley1994;@Seeley2000]. Although bees that follow dances do not specifically interpret this information on an individual level [@Seeley1991], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A]."}
knitr::include_graphics("../Results/figures/manuscript_figures/fig1.png")
```

Despite its complexity, research has shown that the ***value*** collective foraging provides to foraging success is likely to be realized only in limited ecological contexts, the nature of which has proved hard to identify empirically [@Dornhaus1999;@Sherman2002;@Dornhaus2004;@Dornhaus2006;@Okada2012;@Price2019;@Donaldson-Matasci2012;@Donaldson-Matasci2014@Nurnberger2017]. Since dancing foragers use the top of a comb to reference the sun's azimuth [@Frisch1993], it is possible to render dances meaningless by placing combs on a horizontal plane, and thus force foragers to search out food individually, preventing collective decisions. Initial work following these methods tentatively linked the benefits of collective foraging to landscape heterogeneity [@Sherman2002], but empirical attempts to systematically test this hypothesis have failed to provide support [@Donaldson-Matasci2012;@Nurnberger2017], and dance disruption in challenging environments has even sometimes been associated with higher, rather than lower, foraging success [@Price2019]. Such mixed evidence has limited progress in identifying the ecological conditions that originally shaped the evolution of the dance communication system, and thus the collective behaviour of these extraordinary eusocial insects [@Price2015].

Here, we adopt an alternative approach that seeks to identify those environments in which colonies forage collectively and those in which individual search is dominant. It is well-established that that bees do not always choose to dance on return from a foraging trip [@Frisch1993], and furthermore, that individuals often ignore the dance when looking for forage sites [@Gruter2009]. Thus, colonies are likely to vary in their reliance on collective foraging, and identifying the ecological variables that drive this variation offers a means to establish what contexts may have been key to dance evolution. To quantify reliance on the dance within colonies, we adopt a modelling approach that draws on the distribution of foraging distances reported on the dance floor of individual hives, thus avoiding the time -and labour- intensive need to document foragers' individual search histories [@Seeley1995]. Our model provides a description of the distribution of the distances reported on the dance floor and can be used to quantify to what extent colonies rely on collective or individual foraging. We apply our model to waggle dance observations collected from 20 sites across and agri-rural and urban environment (10 in each), and find that some colonies forage almost entirely collectively while others are dominated by individual search. We then use our model to explore how variations in landscape structure influence dance use in colony foraging within our dataset, demonstrating how our methodology could be used to seek out those ecological circumstances under which collective foraging is important.

@@

To test whether colonies that rely on individual search might differ from those that rely on recruitment in terms of the distribution of foraging distributions reported by dancing bees, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor [@Seeley1992] (Fig. 2a,b, details in Materials and Methods). As it is known in the simulation what proportion of individuals in the hive forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor by each type of forager. Fig 2c shows that the shapes of the resource distance distributions for bees engaging in the two types of foraging trips are different. The distance distribution for the scout trips is close to that of an exponential distribution (Fig. 2c), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see Materials and Methods). The distribution of the distances reported for recruit trips (Fig. 2c) is a Rayleigh distribution which is the nearest-neighbour distribution in a two-dimensional environment [@Pyke1978] (see Materials and Methods).

```{r, fig.cap="Figure 2. **Simulating honeybee foraging**. In our simulation model with scouting only (A), foragers leave the hive on a search path (white lines) and continue until they encounter a resource (circles, colours indicate different resource quality). When foraging with recruitment (B) foragers continue to identify resources in scouting trips (white lines) and convey this information on the dance floor (brown disc) where foragers can sample dances reporting on scouting and recruiting trips and follow these directions (yellow lines). (C) Complementary cumulative frequencies of foraging distances reported from scouting and recruit trips. Note the difference in the shape of the distributions. The scout distribution is best fit by an exponential (grey fit line), the recruit distribution by a Rayleigh distribution (black line)"}
knitr::include_graphics("../Results/figures/manuscript_figures/fig2.png")
```

@@

We constructed two models of foraging: an individual model describing a hypothetical colony that relies only on individual search to find foraging sites, and a collective model which describes a colony in which a proportion of bees follow waggle dances to find resources. In the individual model, all trips are "scout" trips, but as recruitment becomes more important, the proportion of “recruit” trips will increase. By setting $p = 1$, we obtain a model based on the sole use of individual search (scouts only) and, alternatively, by allowing the proportion of scout trips, $p$, to take on any value between 0 and 1, we can model the extent to which foraging occurs collectively. (Fig 2). In fitting these two models to distribution of waggle run durations decoded from real honeybee colonies we are thus able, using model selection [@Burnham2002], to infer if and, by estimating the parameter $p$, to what extent honeybee colonies use waggle dance information when foraging.

@@

```{r, fig.cap="Figure 4. **The honeybee foraging model fitted to data from 20 hives**. (A) Location of study hives in Southern England, shaded area in the main plot indicates Greater London. For 16 hives for the collective foraging model provided best explanation (black circles) or for 4 hives the individual search model provided the best explanation (red triangles) as indicated by lowest AIC score. (B). Distribution of goodness of fit confidence values for each model fit to waggle run durations from each site. The p-value is derived from a bootstrapped two-sided KS test comparing the fitted model predictions to the empirical data, the red dashed line marks the significance threshold of 0.05. For values exceeding the threshold there is no statistically significant difference between the model and the data, indicating the model provides a good fit. For the hive in (C) the individual model (blue line) provided a better fit than the collective foraging model (red line). For the hive in (D) the collective foraging model (red line) provided a better fit than the individual model (blue line). The typical “hump” in the distribution in (D) which is indicative of contribution of recruitment dances (compare to Fig 3H). Panels show the compliment cumulative frequencies with binned frequency distributions as inset."}
knitr::include_graphics("../Results/figures/manuscript_figures/fig4.png")
```

Since our model of individual foraging provides a good fit to colony foraging in 11 different sites, and provides a more parsimonious description of foraging than a model of collective foraging in 4 sites, our results indicate that, whilst colony-level foraging is mostly comprised of a mixture of scout and recruit foraging trips, in some circumstances, colony foraging can be better described by individual foraging alone. Thus, in some environments, the majority of foraging trips involve scouting to find new food sites rather than recruitment through dances. Note that this does not imply that these bees do not engage in dance following, because bees regularly follow dances but choose not to visit the advertised site [@Gruter2011], so it may be the case that dances are followed and ignored, not followed, or not performed. However, though the mechanism by which flexibility is achieved at the individual level remains unclear, we have demonstrated that colonies themselves do not always forage collectively.

Quantifying the use of waggle-dance recruitment within all colonies, as a proportion of all foraging trips, can be achieved by extracting the estimated proportion of recruit trips, $1-p$, for each site. Since our sites varied in land-use characteristics, and thus potentially forage distributions (although in this case, not by design for this study), we investigated whether these estimates might correlate with land-use. We first classified the different land-use types of the area surrounding each site [@Samuelson2018] to obtain a standardised land-use profile for the urban and agri-rural environments separately as many land-use types present in urban areas do not occur in agri-rural environments and vice versa (see Materials and Methods). We then performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see Materials and Methods) to determine the principal components that represent combinations of land-use types which explained the most variation in waggle dance use within agri-rural and urban environments. For one of our sites neither model provided a plausible description, so this site was removed from the PLS analysis.

Waggle dance use varied widely regardless of environment, ranging from absolute ($p = 0$) to none ($p = 1$) in both the urban and agri-rural environments. This proportion varied systematically with land-use. In the agri-rural sites, a single principal component explained ~73% of the variation in waggle dance use (beta regression: $R^2 = 0.73$, $\phi$ = 4.9, p < 0.05, Fig. 5A). Similarly, in the urban environment, a single principal component explained ~73% of the variation in waggle dance use (beta regression: $R^2$ = 0.73, $\phi$ = 10.4, p < 0.05, Fig. 5C; note that since the land-use categories differ between the two environments, this single component differs between the two categories). Probing the individual loadings of these principal components allows us to evaluate the driving mechanisms in more detail. As our sample size is limited  (10 urban and 9 agri-rural sites, reduced from 10), we used jackknife resampling to evaluate the robustness of our results to influential points (see Materials and Methods, Supplementary Material) and evaluate how these land-use types contribute to the principal component which significantly correlates with waggle dance use.

@@

In the agri-rural environments the first principal component correlates positively with arable land (29% of land coverage; Table 1) and negatively with built-up areas (17% of land coverage); note that it also correlates negatively with non-agricultural unmanaged green space and water, but together these represent less than 3% of land-use (Table 1; Fig. 5B). These land-use types maintain a significant correlation with the first principal component over the jackknifed PLS (Fig. 5B), with the exception of non-agricultural unmanaged green space which sits on the boarder, indicating the results are robust. As arable land increases whilst built-up areas decrease, the proportion of trips that are driven by individual search increases. Arable land in the UK is typically considered nutritionally poor for bees (note that oilseed rape fields were not included within this category), while there is evidence to suggest that the residential areas that were captured within the “built-up” category are forage-rich hotspots, typically supporting relatively high bee diversity and abundance within gardens [@Baldock2019]. Thus, our agri-rural results suggest that collective foraging, driven by waggle dance recruitment, reduces as food becomes sparser.

In the urban environment the first principal component correlated negatively with the dominant land-use type by coverage, sparse residential (land coverage ~35%; Table 1), since such land includes a high proportion of gardens, this is in keeping with the hypothesis that collective foraging increases in flower-rich areas. However, this relationship is very sensitive to site removal through the jackknife sampling (Supplementary Fig. 2.), and is not a significant contributor to the variance in the loadings (Fig. 5D.). The first principal component also correlates positively with railways but these constitute ~1% of land cover (Table 1). Furthermore, the first principal component correlates negatively with parks, allotments and cemeteries which would typically be considered forage-rich for honeybees [@Baldock2019] (although note that the amount of land cover for this component is limited and the jackknifed PLS identified substantial variation in the loadings identified for the first principal components; Supplementary Fig. 3.). Overall, our findings for the urban environments present further support for a relationship between collective foraging reliance and land-use.

@@

By developing a method to quantify the extent to which a colony forages collectively, here we have shown that colony foraging varies in how much it is dominated by the waggle dance recruitment system.  Furthermore, we have demonstrated how this variation can be compared across land-use types to evaluate how environmental factors shaping collective foraging at the colony level. Whilst recruitment is well known to occur through other mechanisms besides the waggle dance, such as through olfactory cues [@Arenas2007], the exceptionally close fit of our model to waggle run durations underlines the importance of the waggle dance in honeybee foraging.

Within both the agri-rural and urban environments, collective foraging varied with land-use type, ranging from near-total reliance on recruitment to the dominance of individual search. Understanding the circumstances under which colonies forage collectively is a step towards identifying those types of environmental feature that may have made collective foraging beneficial for ancestral honeybees [@Price2015]. Our findings hint that collective foraging may more commonly occur in rich environments, rather than those where food is hard to find. However, this suggestion is speculative for two reasons. Firstly, we used a pre-existing dataset and did not systematically vary landscape traits that may be important, such as forage heterogeneity. We hypothesise that arable land is likely to be more patchy than built-up areas, however, this requires further testing through larger-scale study designed for the purpose.

One of the main outcomes of our study is the development of our methodology to quantifying collective foraging, which can be applied without manipulating colonies through dance disorientation. By combining this tool with further dance-decoding datasets, we envisage a picture should emerge of the circumstances under which honeybee colonies forage collectively. To this end, a major obstacle has recently been overcome thanks to the development automated decoding protocols [@Wario2017;@Wild2021]. Whilst the benefits of dance communication are known to vary across environments [@Donaldson-Matasci2012;@Gruter2011;@Schurch2014;@Price2019], our results show that colonies vary their use in response, affording the colony a flexible dance use strategy which they can tailor to different conditions to maximise foraging success. Identifying the circumstances in which dance communication is important have been hampered by the requirement to monitor real-world colony weight [@Sherman2002;@Dornhaus2004;Gruter2013;@Price2019;@Dornhaus2006;@Nurnberger2017;@Kirchner1998] -a noisy proxy of foraging success that is influenced by many other abiotic and biotic factors, including colony health- over ecologically credible time periods [@Schurch2014]. Doing this at a scale that allows inter-colony variation in foraging environments at sufficient replication is a major logistical hurdle (but see @Nurnberger2017), particularly given that multiple landscape variables may interact to determine the utility of dance communication. Our method circumvents this issue by providing a time and labour efficient methodology to quantify this key collective behaviour and map the environment along the major axis of honeybee information use.

Examining the land-use types which contribute most to the variation in waggle dance use in each of our study environments reveals insights into the drivers of waggle dance use. Our agri-rural results suggest a potential decrease in reliance on waggle dance recruitment as resources become harder to find and foraging trips become longer [@Samuelson2021], broadly agreeing with some of the findings from other theoretical [@Dornhaus2006] and empirical [@Price2019] studies which examined foraging performance with and without the dance, but contrasting with others. These results suggest that, even though collective foraging is not always beneficial, in environments where high-quality resources are present but relatively scarce, collective foraging is being carried out by colonies and under these conditions exchanging social information through the waggle dance confers benefits. In both agri-rural and urban landscapes we have shown that the waggle dance use changes with land-use (illustrated in the change in land-use shown in the maps from left to right in Fig 5A and 5C). These findings illustrate how our model can be used to estimate reliance on collective behaviour within different landscapes, for exploration within datasets where land-use is systematically chosen by design.

Although the number of colonies used in this analysis are comparable to similar studies evaluating collective foraging in honeybees [@Sherman2002;@Dornhaus2004;@Price2019;@Nurnberger2017;@Kirchner1998],  probing the landscape drivers of colony level waggle dance use in more detail requires more colony level data. As longer term effects, such as resource stability [@Schurch2014], may also influence waggle dance use, these data need to be gathered over sufficient time to ensure these effects are accounted for. Here, we have presented an accurate mathematical description of colony foraging which extends our ability to quantify collective behaviour across environments, providing a methodology to assist in identifying the selection pressures that may have driven the evolution of this remarkable behaviour. This will elucidate how plasticity in collective decision making can help to buffer the effects of landscape change, and may help to exploit resources when experiencing novel landscape mosaics, such as in highly urbanized conurbations.

@@

All of the data and code required to reproduce this article can be found at:
